---
title: "Untitled"
author: "Fumin Li"
date: "`r Sys.Date()`"
output: html_document
---

Task1

```{r}
library(cluster)   # for clusGap
library(ggplot2)
library(dplyr)

# 1. Generate clustered data in n-dim hypercube

generate_hypercube_clusters <- function(n, k, side_length, noise_sd = 1.0) {
  # Cluster centers: the "positive corners" of n-d hypercube
  centers <- diag(rep(side_length, n))
  
  # For each cluster, generate k samples around the center
  data_list <- lapply(1:n, function(i) {
    matrix(rnorm(k * n, mean = centers[i, ], sd = noise_sd), ncol = n, byrow = TRUE)
  })
  
  data <- do.call(rbind, data_list)
  labels <- rep(1:n, each = k)
  
  return(list(X = data, y = labels))
}

# 2. Run clusGap for given dataset

estimate_num_clusters <- function(X, maxK = 10) {
  set.seed(123)
  gap <- clusGap(X, FUN = kmeans, nstart = 20, K.max = maxK, B = 50, iter.max = 50)
  bestK <- maxSE(gap$Tab[,"gap"], gap$Tab[,"SE.sim"], method = "Tibs2001SEmax")
  return(list(bestK = bestK, gap = gap))
}

# 3. Simulation loop

run_simulation <- function() {
  k <- 100
  noise_sd <- 1.0
  dims <- c(6, 5, 4, 3, 2)
  side_lengths <- 10:1
  
  results <- data.frame()
  
  for (n in dims) {
    for (L in side_lengths) {
      cat(sprintf("Running n=%d, side_length=%d\n", n, L))
      dat <- generate_hypercube_clusters(n, k, L, noise_sd)
      est <- estimate_num_clusters(dat$X, maxK = 10)
      results <- rbind(results, 
                       data.frame(n = n, side_length = L, 
                                  est_clusters = est$bestK, true_clusters = n))
    }
  }
  return(results)
}

# 4. Visualization

plot_results <- function(results) {
  ggplot(results, aes(x = side_length, y = est_clusters, color = factor(n))) +
    geom_line(size = 1) +
    geom_point(size = 2) +
    geom_hline(aes(yintercept = true_clusters), linetype = "dashed", color = "black") +
    scale_x_reverse(breaks = 10:1) +
    labs(
      title = "Estimated Number of Clusters vs. Side Length",
      x = "Side length (clusters closer → smaller)",
      y = "Estimated number of clusters (Gap Statistic)",
      color = "Dimension (n)"
    ) +
    theme_minimal(base_size = 14)
}

# 5. Run 

results <- run_simulation()
print(results)
plot_results(results)

```

The plot shows that for all tested dimensions (n = 2–6), the Gap Statistic correctly identifies the true number of clusters when the side length is large (>=4). As the clusters move closer (side length < 4), the estimated number of clusters sharply drops to 1, indicating that the algorithm can no longer separate them. Therefore, the critical transition occurs around side length 3–4, below which cluster distinction becomes unreliable.

Task2
```{r}


library(cluster)
library(ggplot2)
library(plotly)
library(Matrix)


# 1. Generate concentric shells in 3D

generate_shell_clusters <- function(n_shells, k_per_shell, max_radius, noise_sd = 0.1) {

  radii <- seq(1, max_radius, length.out = n_shells)
  data <- NULL
  labels <- NULL
  
  for (i in seq_along(radii)) {
    r <- radii[i]
    theta <- runif(k_per_shell, 0, 2 * pi)
    phi <- acos(runif(k_per_shell, -1, 1))
    
    x <- r * sin(phi) * cos(theta)
    y <- r * sin(phi) * sin(theta)
    z <- r * cos(phi)
    
    shell <- cbind(x, y, z) + matrix(rnorm(3 * k_per_shell, 0, noise_sd), ncol = 3)
    data <- rbind(data, shell)
    labels <- c(labels, rep(i, k_per_shell))
  }
  
  return(list(X = data, y = labels))
}


# 2. Spectral Clustering Function

spectral_clustering <- function(X, k, d_threshold = 1) {
  n <- nrow(X)
  
  dist_mat <- as.matrix(dist(X))
  A <- matrix(0, n, n)
  A[dist_mat < d_threshold] <- 1
  diag(A) <- 0

  D <- diag(rowSums(A))
  D_inv_sqrt <- diag(1 / sqrt(diag(D) + 1e-8))
  L_sym <- diag(n) - D_inv_sqrt %*% A %*% D_inv_sqrt

  eig <- eigen(L_sym, symmetric = TRUE)
  eigvecs <- eig$vectors[, (n - k + 1):n] 

  set.seed(123)
  km <- kmeans(eigvecs, centers = k, nstart = 20, iter.max = 50)
  return(km$cluster)
}


# 3. Wrapper for clusGap compatibility

spectral_wrapper <- function(x, k, ...) {
  labels <- spectral_clustering(x, k, d_threshold = 1)
  
  centers <- aggregate(x, by = list(labels), mean)[, -1]
  wss <- sum(sapply(unique(labels), function(c) {
    sum(rowSums((x[labels == c, ] - centers[c, ])^2))
  }))
  
  return(list(cluster = labels, tot.withinss = wss))
}
# 4. Simulation over radii
run_simulation_shells <- function() {
  n_shells <- 4
  k_per_shell <- 100
  noise_sd <- 0.1
  d_threshold <- 1
  radii <- 10:1
  
  results <- data.frame()
  
  for (r in radii) {
    cat(sprintf("Running spectral clustering with max_radius = %d\n", r))
    dat <- generate_shell_clusters(n_shells, k_per_shell, r, noise_sd)
    
    gap <- clusGap(dat$X, FUN = spectral_wrapper, nstart = 1, K.max = 10, B = 30)
    bestK <- maxSE(gap$Tab[, "gap"], gap$Tab[, "SE.sim"], method = "Tibs2001SEmax")
    
    results <- rbind(results, data.frame(max_radius = r, est_clusters = bestK, true_clusters = n_shells))
  }
  
  return(results)
}

# 5. Visualization
plot_results_shells <- function(results) {
  ggplot(results, aes(x = max_radius, y = est_clusters)) +
    geom_line(size = 1, color = "#0072B2") +
    geom_point(size = 2, color = "#E69F00") +
    geom_hline(aes(yintercept = true_clusters), linetype = "dashed", color = "black") +
    scale_x_reverse(breaks = 10:1) +
    labs(
      title = "Spectral Clustering: Estimated Number of Shells vs Max Radius",
      x = "Maximum Radius (shells closer → smaller)",
      y = "Estimated number of clusters"
    ) +
    theme_minimal(base_size = 14)
}

# 6. 3D visualization for one dataset
plot_shells_3D <- function(X, labels) {
  plot_ly(
    x = X[,1], y = X[,2], z = X[,3],
    color = as.factor(labels),
    type = "scatter3d", mode = "markers",
    marker = list(size = 3)
  )
}


# 7.
results_shells <- run_simulation_shells()
print(results_shells)
plot_results_shells(results_shells)

```

The plot shows that the spectral clustering algorithm can correctly identify four shells only when the maximum radius is large enough for clear separation between shells. When the maximum radius becomes smaller than about seven, the estimated number of clusters quickly drops to one or two, meaning the shells are no longer distinguishable because they overlap. The small fluctuations at larger radii indicate minor instability in the eigenvalue step when the shells are close to the distance threshold. Overall, spectral clustering begins to fail once the shells are too close compared with the fixed distance threshold of one.
